---
title: "CSC 308 Final Project - Individual Portfolio"
author: "Chloe"
date: "April 2026"
output: html_document
---
  
  ```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
```

# Introduction

For this project, my role focused on **discussing the implications of the results and explaining why the findings matter**.  
Instead of mainly creating the exploratory analysis, my contribution centered on interpreting the significance of the patterns found in the data and connecting them to real-world healthcare access issues.

The purpose of my analysis was to explain how the project’s findings relate to healthcare inequality, differences across communities, and the broader importance of socioeconomic and health-related conditions.

---
  
  # Dataset and Preparation
  
  I used the cleaned 2021 project dataset:
  
  `processed_data/2021/Master_2021_Cleaned.csv`

The analysis focused on the following states:
  
  - Virginia  
- North Carolina  
- Maryland  
- Georgia  
- Florida  

```{r}
data21 <- read.csv("processed_data/2021/Master_2021_Cleaned.csv")

states <- c("Virginia", "North Carolina", "Maryland", "Georgia", "Florida")

filtered_data <- data21 %>%
  filter(State %in% states)

head(filtered_data)
```

---
  
  # Summary Statistics
  
  I first reviewed summary statistics to support discussion of the project findings.

```{r}
summary_table <- filtered_data %>%
  group_by(State) %>%
  summarise(
    avg_income = mean(Median.Household.Income, na.rm = TRUE),
    avg_clinical_rank = mean(Clinical_Care_Rank, na.rm = TRUE)
  )

summary_table
```

## Discussion

This summary helps show that the selected states differ in both income and clinical care rankings.  
These differences matter because they suggest that economic conditions may shape the quality of healthcare access available in different communities.

The results also show that healthcare disparities are not evenly distributed. Some states appear to have stronger average conditions than others, which may reflect broader differences in resources, provider availability, and community health support.

---
  
  # Visualization: Income vs Clinical Care
  
  ```{r}
ggplot(filtered_data, aes(x = Median.Household.Income, y = Clinical_Care_Rank)) +
  geom_point(alpha = 0.4) +
  labs(
    title = "Income vs Clinical Care (2021)",
    x = "Median Household Income",
    y = "Clinical Care Rank"
  ) +
  theme_minimal()
```

## Discussion

This plot suggests that counties with higher household income tend to have better clinical care rankings.  
Because lower ranking values indicate better clinical care, the pattern implies that wealthier counties may have advantages in healthcare access and quality.

This has important implications. It suggests that healthcare access is connected to economic inequality. Communities with fewer financial resources may also face greater difficulty accessing providers, services, and other forms of care.

In a real-world sense, this means healthcare differences are not only medical issues. They are also social and economic issues.

---
  
  # Correlation: Income vs Clinical Care
  
  ```{r}
correlation <- cor(
  filtered_data$Median.Household.Income,
  filtered_data$Clinical_Care_Rank,
  use = "complete.obs"
)

correlation
```

## Discussion

The correlation is negative, which supports the idea that higher income is associated with better clinical care rankings.

This result matters because it strengthens the visual pattern shown in the scatter plot.  
It suggests that income may play a meaningful role in shaping healthcare access across counties.

The implication of this finding is that improving healthcare equity may require more than expanding medical services alone. It may also require attention to larger economic inequalities.

---
  
  # Visualization: Average Income by State
  
  ```{r}
ggplot(summary_table, aes(x = State, y = avg_income)) +
  geom_col() +
  labs(
    title = "Average Income by State (2021)",
    x = "State",
    y = "Average Income"
  ) +
  theme_minimal()
```

## Discussion

This state comparison shows that average income is not the same across the selected states.  
Some states have noticeably higher average income than others.

This is important because differences in income may help explain why healthcare rankings are also uneven.  
If income shapes access to insurance, transportation, providers, and preventive care, then economic differences may contribute to long-term health disparities across regions.

---
  
  # Correlation: Health Factors vs Health Outcomes
  
  ```{r}
correlation2 <- cor(
  filtered_data$Health_Factors_Rank,
  filtered_data$Health_Outcomes_Rank,
  use = "complete.obs"
)

correlation2
```

## Discussion

The relationship between health factors and health outcomes is very strong.  
This means counties with weaker health-related conditions also tend to have poorer overall health outcomes.

This finding has a broader implication for public health. It suggests that healthcare outcomes are shaped by more than treatment alone. Social and environmental conditions also matter.

Because of this, policies that focus only on hospitals or medical care may not be enough. Improving health outcomes may also require investments in the larger conditions that affect people’s daily lives.

---
  
  # Real-World Implications
  
  The results of this project suggest several important implications:
  
  - Economic inequality may contribute to unequal healthcare access  
- Better clinical care is often connected to stronger financial conditions  
- Health outcomes are influenced by broader health factors, not only medical care  
- Communities with weaker support conditions may face long-term disadvantages in health and healthcare access  

These findings matter because they show that healthcare disparities are part of a larger structural issue.  
This means that solving healthcare inequality likely requires both medical improvements and attention to larger social and economic conditions.

---
  
  # Limitations
  
  Although the results are meaningful, there are still some limitations.

First, rankings and summary values do not capture every detail about the healthcare experience of individuals.  
Second, correlation does not prove causation. Even though the relationships are strong, the data does not prove that one variable directly causes the other.  
Finally, the analysis focuses on selected states and one cleaned dataset, so the findings should be interpreted within that scope.

---
  
  # Conclusion
  
  My contribution to this project focused on discussing the meaning and significance of the results.  
The analysis suggests that differences in income, clinical care, and health-related conditions are closely connected.

Overall, these results show that healthcare access disparities are tied to larger patterns of inequality.  
This makes the findings important not only for data analysis, but also for understanding real-world healthcare challenges.

---
  
  # Individual Contribution
  
  For this project, I was responsible for:
  
  - Explaining the significance of the project results  
- Discussing the implications of relationships found in the data  
- Connecting findings to real-world healthcare inequality  
- Supporting discussion with summary statistics, visualizations, and correlations  
- Reflecting on limitations and broader meaning of the results  

All code in this document reflects my individual contribution to the project.