---
title: "Healthcare Access Trends (2021–2023)"
subtitle: "Implications of Results, Analytical Discussion, and Interpretation"
author: "Chloe Washington"

output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 6
)
```

---

## Project Overview

**Course:** CSC 308 — Organization of Programming Languages  
**Project:** Healthcare Access Trends (2021–2023)  
**Role:** Implications of Results, Analytical Discussion, and Interpretation

This report supports the analytical discussion component of the group project. The goal is to examine what the project's findings mean in a real-world healthcare context — specifically, how income disparities across U.S. counties relate to clinical care quality, and how behavioral and environmental health factors connect to population health outcomes.

---

## Setup

Load the required libraries for data manipulation and visualization.

```{r libraries}
library(dplyr)
library(ggplot2)
```

---

## Section 1: Load the Dataset (2021)

We begin by loading the cleaned 2021 dataset created earlier in the project pipeline.

```{r load-data}
# Use the cleaned dataset created earlier in the project
data21 <- master_2021

# Quick check that the data loaded correctly
head(data21)
```

---

## Section 2: Use Full Dataset (All U.S. Counties)

No state-level filtering is applied here. The analysis reflects all U.S. counties to ensure the findings are nationally representative.

```{r full-dataset}
# No filtering applied — analysis reflects all states
filtered_data <- data21
```

---

## Section 3: Summary Table by State

The table below shows the average median household income and average clinical care rank for each state. This provides a high-level view of regional variation in both economic conditions and healthcare quality.

```{r summary-table}
summary_table <- filtered_data %>%
  group_by(State) %>%
  summarise(
    avg_income        = mean(Median.Household.Income, na.rm = TRUE),
    avg_clinical_rank = mean(Clinical_Care_Rank, na.rm = TRUE)
  )

summary_table
```

> **Interpretation:** Differences in average income across states highlight regional disparities that may contribute to unequal healthcare access and outcomes. States with lower average incomes tend to cluster toward worse clinical care rankings, suggesting a systemic relationship between economic resources and healthcare quality.

---

## Section 4: Income vs. Clinical Care Rank (Scatter Plot)

The scatter plot below visualizes the relationship between median household income and clinical care rank at the county level across the United States in 2021.

```{r scatter-plot}
ggplot(filtered_data, aes(x = Median.Household.Income, y = Clinical_Care_Rank)) +
  geom_point(alpha = 0.4, color = "#2C7BB6") +
  labs(
    title    = "Income vs. Clinical Care Across U.S. Counties (2021)",
    subtitle = "Each point represents one U.S. county",
    x        = "Median Household Income",
    y        = "Clinical Care Rank (Lower = Better)"
  ) +
  theme_minimal(base_size = 13)
```

> **Interpretation:** The scatter plot reveals a downward trend — as median household income increases, clinical care rank tends to decrease (i.e., improve). This pattern points to a meaningful association between a county's economic standing and the quality of clinical care available to its residents.

---

## Section 5: Correlation — Income and Clinical Care

To quantify the relationship observed in the scatter plot, we compute the Pearson correlation coefficient between median household income and clinical care rank.

```{r correlation-income-care}
correlation <- cor(
  filtered_data$Median.Household.Income,
  filtered_data$Clinical_Care_Rank,
  use = "complete.obs"
)

cat("Correlation between income and clinical care rank:", round(correlation, 4))
```

> **Interpretation:** The negative correlation confirms that counties with higher incomes tend to have better (lower-numbered) clinical care rankings. This suggests that economic resources may play a significant role in shaping access to and quality of healthcare services. Counties with fewer financial resources may face systemic barriers — such as fewer healthcare providers, reduced insurance coverage, and limited infrastructure — that translate directly into worse care outcomes.

---

## Section 6: Average Income by State (Bar Chart)

The bar chart below ranks all U.S. states by their average county-level median household income, providing a geographic view of economic inequality.

```{r bar-chart-income}
ggplot(summary_table, aes(x = reorder(State, avg_income), y = avg_income)) +
  geom_col(fill = "#4DAF4A") +
  coord_flip() +
  labs(
    title    = "Average Median Household Income by State (2021)",
    subtitle = "Based on county-level averages",
    x        = "State",
    y        = "Average Income ($)"
  ) +
  theme_minimal(base_size = 11)
```

> **Interpretation:** The bar chart makes visible the substantial income gap between states. States at the top of the chart — with higher average incomes — are generally located in the Northeast and West Coast, while lower-income states are concentrated in the South and rural Midwest. These geographic patterns mirror known disparities in healthcare infrastructure and outcomes, reinforcing the implication that place of residence is a strong predictor of healthcare access.

---

## Section 7: Health Factors vs. Health Outcomes

Finally, we examine whether a county's overall health factors (behavioral, environmental, and social determinants) are associated with its health outcomes (length and quality of life).

```{r correlation-health}
correlation2 <- cor(
  filtered_data$Health_Factors_Rank,
  filtered_data$Health_Outcomes_Rank,
  use = "complete.obs"
)

cat("Correlation between health factors and health outcomes rank:", round(correlation2, 4))
```

> **Interpretation:** The strong positive correlation between health factors and health outcomes ranks indicates that counties with poorer underlying health conditions — such as high rates of smoking, obesity, or physical inactivity, combined with limited social and economic resources — also tend to experience worse overall health outcomes. This finding underscores the importance of upstream, preventive interventions that address root causes of poor health rather than focusing solely on clinical treatment after illness occurs.

---

## Conclusion

The analyses presented in this report consistently point to a central theme: **economic and environmental conditions are deeply intertwined with healthcare access and population health**. Key takeaways include:

- Counties and states with higher median household incomes tend to rank better in clinical care quality.
- Geographic disparities in income map closely onto disparities in healthcare access, suggesting that place of birth and residence remains a major determinant of health equity.
- Health factors — encompassing behavioral, environmental, and socioeconomic influences — are strongly correlated with health outcomes, reinforcing the case for holistic, community-level health policy interventions.

These findings have real policy implications: improving healthcare access in underserved areas may require addressing income inequality, expanding insurance coverage, investing in rural and low-income healthcare infrastructure, and targeting public health programs at the communities with the greatest need.

---

*CSC 308 — Organization of Programming Languages | Final Project*
