---
Name: Mady Boyd, Emmanuella Obidike, Chloe Washingotn
title: "R Notebook"
output: html_notebook
---
Access to healthcare is not the same for everyone. Differences often depend on income, where people live, and demographic factors. Although healthcare data is widely available, it can be hard to understand without proper analysis and visualization. Because of this difficulty, people are not able to clearly see where the gaps and differences in access to healthcare exist and what factors could be the potential causes for those gaps. The purpose of this project is to use a form of data analysis and visualization to break down and understand patterns regarding healthcare access and point out those disparities across different demographics and populations. With this project, complex healthcare data will be easier to interpret and understand for people like students, researchers, and all.

This project aims to explore healthcare access disparities through a data-focused and visualization-centered approach using the R programming language. By analyzing measurable factors such as income, insurance coverage, geographic location, and demographic characteristics, the project seeks to make complex healthcare data more accessible and understandable to students, researchers, and other stakeholders.
---
title: "CSC 308 Final Project - Individual Portfolio"
author: "Mady"
date: "April 2026"
output: html_document
---

4.1 Data Cleaning and Processing Logic (2021) 

The 2021 analytical baseline was built using a modular cleaning pipeline. Since the County Health Rankings data is stored in specialized files, the method required a "Deconstruction and Extraction" approach to separate the metrics that are important for the project. 

4.1.1 Ingestion of Socioeconomic Predictors (Files 1 & 2) 

The first step involved processing the "Additional Measure Data" along with the metadata for this file. The main challenge was the presence of metadata headers in the raw CSV. This required using the skip = 1 argument to match the headers with the data rows. 

```{r clean_2021_file1}
#Processing File 1: Additional Measure Data 
# Standardizing 'null' values to handle suppressed data points

 clean21_measures <- read.csv("data/raw/2021/Additional Measure Data.csv", skip = 1, na.strings = c("N/A", "", " ")) %>% dplyr::select(FIPS, State, County, Median.Household.Income, X..Uninsured, X..Rural) %>% filter(County != "") 

# Verification: Ensure Socioeconomic columns are present names(clean21_measures) 
```r


4.1.2 Normalizing Health Rankings (Files 3 & 4) 

Files 3 (Outcomes and Factors Rankings) and 4 (Outcomes and Factors SubRankings) gave the "calculated results" of the study. Because the raw files used generic headers like Rank and Rank.1, I used a Descriptive Renaming Strategy during selection to avoid variable collision when creating the final master join. 

```{r clean_2021_file3_4}
# Processing File 3: Core Rankings 
clean21_ranked <- read.csv("data/raw/2021/Outcomes and Factors Rankings.csv", skip = 1) %>% dplyr::select(FIPS, Health_Outcomes_Rank = Rank, Health_Factors_Rank = Rank.1) 

# Processing File 4: Specialized Sub-Rankings 

clean21_subranked <- read.csv("data/raw/2021/Outcomes and Factors SubRankings.csv", skip = 1) %>% dplyr::select(FIPS, Length_of_Life_Rank = Rank, Clinical_Care_Rank = Rank.3, Socioeconomic_Rank = Rank.4)
```r

 4.1.3 Isolating Racial Disparity Metrics (Files 5 & 6) 

File 5 (Ranked Measure Data) had over 100 columns, including raw rates, confidence intervals, and data reliability flags. To keep the focus on racial equity, I used the contains() function to target the main Years of Potential Life Lost (YPLL) rates for Black and White populations while discarding secondary statistics. 

```{r clean_2021_file5}
 # Processing File 5: Targeted extraction of YPLL and Poor Health metrics clean21_rankedmeasures <- read.csv("data/raw/2021/Ranked Measure Data.csv", skip = 1) %>% dplyr::select(FIPS, YPLL_Rate = contains("Potential.Life.Lost.Rate"), YPLL_Black = contains("YPLL.Rate..Black"), YPLL_White = contains("YPLL.Rate..White"), Percent_Poor_Health = contains("Fair.or.Poor.Health")) %>% filter(!is.na(YPLL_Rate)) 

```r

4.1.4 Relational Integration and Data Type Audit 

The last step of the 2021 method was a multi-stage Relational Join using the Federal Information Processing Standard (FIPS) codes as the main key. After the join, I did a Data Type Audit. Because raw CSVs often import numeric rankings as character strings due to alphanumeric placeholders like "NR" (Not Reported), I converted these to numeric doubles. 


# Synthesizing the 2021 Master Dataset 

```{r join_and_audit_2021}

master_2021 <- clean21_measures %>% inner_join(clean21_ranked, by = "FIPS") %>% inner_join(clean21_subranked, by = "FIPS") %>% inner_join(clean21_rankedmeasures, by = "FIPS") # Auditing and forcing numeric types for mathematical validity master_2021$Median.Household.Income <- as.numeric(as.character(master_2021$Median.Household.Income)) master_2021$Health_Factors_Rank <- as.numeric(as.character(master_2021$Health_Factors_Rank)) master_2021$YPLL_Black <- as.numeric(as.character(master_2021$YPLL_Black)) master_2021$YPLL_White <- as.numeric(as.character(master_2021$YPLL_White)) # Listwise deletion of incomplete records master_2021 <- na.omit(master_2021) # Final Verification print(paste("Final 2021 County Count:", nrow(master_2021))) str(master_2021[c("Median.Household.Income", "YPLL_Black")])
```r
---
title: "CSC 308 Final Project - Individual Contribution"
author: "Emmanuella Obidike"
date: "April 2026"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
```

# Project Overview

This document represents my individual contribution to the group project titled Exploring Healthcare Access Disparities Through Data-Driven Visualization and Analysis. My work focuses specifically on exploratory data analysis and interpretation using the cleaned 2021 dataset.

This project examines relationships between income, healthcare access, and health outcomes using county-level data. The analysis focuses on identifying patterns between economic factors and overall health rankings across selected states.

The dataset includes measures such as median household income, clinical care rankings, and health factors/outcomes indicators. The goal of the project is to understand how socioeconomic conditions influence healthcare quality and population health.

---

# Individual Contribution - Emmanuella Obidike

**Role:** Exploratory Data Analysis (EDA), Statistical Analysis, and Interpretation

For this project, I was responsible for analyzing the cleaned 2021 dataset to explore relationships between income, healthcare access, and health outcomes. My work included filtering the dataset, generating summary statistics, creating visualizations, performing correlation analysis, and interpreting key patterns observed in the data.

All analysis and code presented in this document reflect my individual contribution to the overall team project.

---

# Dataset and Preparation

I used the cleaned dataset `Master_2021_Cleaned.csv`, which contains county-level data including:

- Median household income  
- Clinical care rankings  
- Health factors and health outcomes rankings  
- Demographic and health-related variables  

To stay consistent with the project focus, I filtered the dataset to the following states:

- Virginia  
- North Carolina  
- Maryland  
- Georgia  
- Florida  

```{r}
# Load dataset and filter selected states
data21 <- read.csv("processed_data/2021/Master_2021_Cleaned.csv")
states <- c("Virginia", "North Carolina", "Maryland", "Georgia", "Florida")

filtered_data <- data21 %>%
  filter(State %in% states)


```

The analysis focuses on these states to provide a regional comparison of healthcare access and socioeconomic differences across diverse geographic areas.

---

# Summary Statistics

```{r}
summary_table <- filtered_data %>%
  group_by(State) %>%
  summarise(
    avg_income = mean(Median.Household.Income, na.rm = TRUE),
    avg_clinical_rank = mean(Clinical_Care_Rank, na.rm = TRUE)
  )

summary_table
```

This provides a quick comparison of income and clinical care differences across states

---

# Visualization 1: Income vs Clinical Care

```{r}
ggplot(filtered_data, aes(x = Median.Household.Income, y = Clinical_Care_Rank)) +
  geom_point(color = "blue", alpha = 0.4) +
  labs(
    title = "Income vs Clinical Care (2021)",
    x = "Median Household Income",
    y = "Clinical Care Rank"
  ) +
  theme_minimal()
```

**Interpretation**

This scatter plot shows a clear negative relationship between income and clinical care ranking. 
As income increases, clinical care rank tends to improve (lower rank = better care).

---

# Visualization 2: Average Income by State

```{r}
ggplot(summary_table, aes(x = State, y = avg_income)) +
  geom_col(fill = "purple") +
  labs(
    title = "Average Income by State (2021)",
    x = "State",
    y = "Average Income"
  ) +
  theme_minimal()
```

**Interpretation**

Maryland exhibits the highest average income among the selected states, while Georgia and North Carolina show comparatively lower income levels.

---

# Correlation: Income vs Clinical Care

```{r}
correlation <- cor(
  filtered_data$Median.Household.Income,
  filtered_data$Clinical_Care_Rank,
  use = "complete.obs"
)

correlation
```

**Interpretation**

The correlation is about -0.48, showing a moderate negative relationship, indicating that higher income is associated with better (lower) clinical care rankings.

---

# Visualization 3: Health Factors vs Health Outcomes

```{r}
ggplot(filtered_data, aes(x = Health_Factors_Rank, y = Health_Outcomes_Rank)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  labs(
    title = "Health Factors vs Health Outcomes",
    x = "Health Factors Rank",
    y = "Health Outcomes Rank"
  ) +
  theme_minimal()
```

**Interpretation**

This suggests that counties with poorer health factors tend to have worse health outcomes, indicating a strong relationship between environmental conditions and overall population health.

---

# Correlation: Health Factors vs Outcomes

```{r}
correlation2 <- cor(
  filtered_data$Health_Factors_Rank,
  filtered_data$Health_Outcomes_Rank,
  use = "complete.obs"
)

correlation2
```

**Interpretation**

The correlation is about 0.91, which is very strong.

---

# Conclusion

The analysis reveals clear relationships between socioeconomic factors and healthcare access.

Higher income levels are generally associated with better clinical care rankings, suggesting that economic stability plays a key role in access to healthcare resources. Also, the strong relationship between health factors and health outcomes indicates that environmental and behavioral conditions significantly influence overall population health.

Differences observed across states highlight how geographic and demographic variations contribute to disparities in healthcare access. These findings emphasize the importance of addressing socioeconomic inequalities in order to improve healthcare accessibility and outcomes.

Overall, this project shows how exploratory analysis can help turn complex healthcare data into insights that are easier to understand.  

These findings directly support the broader project goal of identifying disparities in healthcare access and understanding how socioeconomic factors contribute to differences in health outcomes.

---
# Final Statement
All code in this document reflects my individual contribution.

---
title: "CSC 308 Final Project - Individual Portfolio"
author: "Chloe"
date: "April 2026"
output: html_document
---
  
  ```
{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
```

# Introduction

For this project, my role focused on **discussing the implications of the results and explaining why the findings matter**.  
Instead of mainly creating the exploratory analysis, my contribution centered on interpreting the significance of the patterns found in the data and connecting them to real-world healthcare access issues.

The purpose of my analysis was to explain how the project’s findings relate to healthcare inequality, differences across communities, and the broader importance of socioeconomic and health-related conditions.

---
  
  # Dataset and Preparation
  
  I used the cleaned 2021 project dataset:
  
  `processed_data/2021/Master_2021_Cleaned.csv`

The analysis focused on the following states:
  
  - Virginia  
- North Carolina  
- Maryland  
- Georgia  
- Florida  

```{r}
data21 <- read.csv("processed_data/2021/Master_2021_Cleaned.csv")

states <- c("Virginia", "North Carolina", "Maryland", "Georgia", "Florida")

filtered_data <- data21 %>%
  filter(State %in% states)

head(filtered_data)
```

---
  
  # Summary Statistics
  
  I first reviewed summary statistics to support discussion of the project findings.

```{r}
summary_table <- filtered_data %>%
  group_by(State) %>%
  summarise(
    avg_income = mean(Median.Household.Income, na.rm = TRUE),
    avg_clinical_rank = mean(Clinical_Care_Rank, na.rm = TRUE)
  )

summary_table
```

## Discussion

This summary helps show that the selected states differ in both income and clinical care rankings.  
These differences matter because they suggest that economic conditions may shape the quality of healthcare access available in different communities.

The results also show that healthcare disparities are not evenly distributed. Some states appear to have stronger average conditions than others, which may reflect broader differences in resources, provider availability, and community health support.

---
  
  # Visualization: Income vs Clinical Care
  
  ```{r}
ggplot(filtered_data, aes(x = Median.Household.Income, y = Clinical_Care_Rank)) +
  geom_point(alpha = 0.4) +
  labs(
    title = "Income vs Clinical Care (2021)",
    x = "Median Household Income",
    y = "Clinical Care Rank"
  ) +
  theme_minimal()
```

## Discussion

This plot suggests that counties with higher household income tend to have better clinical care rankings.  
Because lower ranking values indicate better clinical care, the pattern implies that wealthier counties may have advantages in healthcare access and quality.

This has important implications. It suggests that healthcare access is connected to economic inequality. Communities with fewer financial resources may also face greater difficulty accessing providers, services, and other forms of care.

In a real-world sense, this means healthcare differences are not only medical issues. They are also social and economic issues.

---
  
  # Correlation: Income vs Clinical Care
  
  ```{r}
correlation <- cor(
  filtered_data$Median.Household.Income,
  filtered_data$Clinical_Care_Rank,
  use = "complete.obs"
)

correlation
```

## Discussion

The correlation is negative, which supports the idea that higher income is associated with better clinical care rankings.

This result matters because it strengthens the visual pattern shown in the scatter plot.  
It suggests that income may play a meaningful role in shaping healthcare access across counties.

The implication of this finding is that improving healthcare equity may require more than expanding medical services alone. It may also require attention to larger economic inequalities.

---
  
  # Visualization: Average Income by State
  
  ```{r}
ggplot(summary_table, aes(x = State, y = avg_income)) +
  geom_col() +
  labs(
    title = "Average Income by State (2021)",
    x = "State",
    y = "Average Income"
  ) +
  theme_minimal()
```

## Discussion

This state comparison shows that average income is not the same across the selected states.  
Some states have noticeably higher average income than others.

This is important because differences in income may help explain why healthcare rankings are also uneven.  
If income shapes access to insurance, transportation, providers, and preventive care, then economic differences may contribute to long-term health disparities across regions.

---
  
  # Correlation: Health Factors vs Health Outcomes
  
  ```{r}
correlation2 <- cor(
  filtered_data$Health_Factors_Rank,
  filtered_data$Health_Outcomes_Rank,
  use = "complete.obs"
)

correlation2
```

## Discussion

The relationship between health factors and health outcomes is very strong.  
This means counties with weaker health-related conditions also tend to have poorer overall health outcomes.

This finding has a broader implication for public health. It suggests that healthcare outcomes are shaped by more than treatment alone. Social and environmental conditions also matter.

Because of this, policies that focus only on hospitals or medical care may not be enough. Improving health outcomes may also require investments in the larger conditions that affect people’s daily lives.

---
  
  # Real-World Implications
  
  The results of this project suggest several important implications:
  
  - Economic inequality may contribute to unequal healthcare access  
- Better clinical care is often connected to stronger financial conditions  
- Health outcomes are influenced by broader health factors, not only medical care  
- Communities with weaker support conditions may face long-term disadvantages in health and healthcare access  

These findings matter because they show that healthcare disparities are part of a larger structural issue.  
This means that solving healthcare inequality likely requires both medical improvements and attention to larger social and economic conditions.

---
  
  # Limitations
  
  Although the results are meaningful, there are still some limitations.

First, rankings and summary values do not capture every detail about the healthcare experience of individuals.  
Second, correlation does not prove causation. Even though the relationships are strong, the data does not prove that one variable directly causes the other.  
Finally, the analysis focuses on selected states and one cleaned dataset, so the findings should be interpreted within that scope.

---
  
  # Conclusion
  
  My contribution to this project focused on discussing the meaning and significance of the results.  
The analysis suggests that differences in income, clinical care, and health-related conditions are closely connected.

Overall, these results show that healthcare access disparities are tied to larger patterns of inequality.  
This makes the findings important not only for data analysis, but also for understanding real-world healthcare challenges.

---
  
  # Individual Contribution
  
  For this project, I was responsible for:
  
  - Explaining the significance of the project results  
- Discussing the implications of relationships found in the data  
- Connecting findings to real-world healthcare inequality  
- Supporting discussion with summary statistics, visualizations, and correlations  
- Reflecting on limitations and broader meaning of the results  

All code in this document reflects my individual contribution to the project.

---
title: "CSC 308 Final Project - Individual Portfolio"
author: "Joshua"
date: "April 2026"
output: html_document
---
This visualization explores the relationship between income and healthcare access. The results will show that lower income levels are associated with higher uninsured rates which portray that economic factors play a vital role in healthcare accessibility.

This analysis supports the overall findings of the project because it shows how financial disparities influence health outcomes across different populations.

