---
Name: Mady Boyd, Emmanuella Obidike, Chloe Washington
title: "R Notebook"
output: html_notebook
---
Access to healthcare is not the same for everyone. Differences often depend on income, where people live, and demographic factors. Although healthcare data is widely available, it can be hard to understand without proper analysis and visualization. Because of this difficulty, people are not able to clearly see where the gaps and differences in access to healthcare exist and what factors could be the potential causes for those gaps. The purpose of this project is to use a form of data analysis and visualization to break down and understand patterns regarding healthcare access and point out those disparities across different demographics and populations. With this project, complex healthcare data will be easier to interpret and understand for people like students, researchers, and all.

This project aims to explore healthcare access disparities through a data-focused and visualization-centered approach using the R programming language. By analyzing measurable factors such as income, insurance coverage, geographic location, and demographic characteristics, the project seeks to make complex healthcare data more accessible and understandable to students, researchers, and other stakeholders.
---
title: "CSC 308 Final Project - Individual Portfolio"
author: "Mady"
date: "April 2026"
output: html_document
---

4.1 Data Cleaning and Processing Logic (2021) 

The 2021 analytical baseline was built using a modular cleaning pipeline. Since the County Health Rankings data is stored in specialized files, the method required a "Deconstruction and Extraction" approach to separate the metrics that are important for the project. 

4.1.1 Ingestion of Socioeconomic Predictors (Files 1 & 2) 

The first step involved processing the "Additional Measure Data" along with the metadata for this file. The main challenge was the presence of metadata headers in the raw CSV. This required using the skip = 1 argument to match the headers with the data rows. 

```{r clean_2021_file1}
#Processing File 1: Additional Measure Data 
# Standardizing 'null' values to handle suppressed data points

 clean21_measures <- read.csv("data/raw/2021/Additional Measure Data.csv", skip = 1, na.strings = c("N/A", "", " ")) %>% dplyr::select(FIPS, State, County, Median.Household.Income, X..Uninsured, X..Rural) %>% filter(County != "") 

# Verification: Ensure Socioeconomic columns are present names(clean21_measures) 
```r


4.1.2 Normalizing Health Rankings (Files 3 & 4) 

Files 3 (Outcomes and Factors Rankings) and 4 (Outcomes and Factors SubRankings) gave the "calculated results" of the study. Because the raw files used generic headers like Rank and Rank.1, I used a Descriptive Renaming Strategy during selection to avoid variable collision when creating the final master join. 

```{r clean_2021_file3_4}
# Processing File 3: Core Rankings 
clean21_ranked <- read.csv("data/raw/2021/Outcomes and Factors Rankings.csv", skip = 1) %>% dplyr::select(FIPS, Health_Outcomes_Rank = Rank, Health_Factors_Rank = Rank.1) 

# Processing File 4: Specialized Sub-Rankings 

clean21_subranked <- read.csv("data/raw/2021/Outcomes and Factors SubRankings.csv", skip = 1) %>% dplyr::select(FIPS, Length_of_Life_Rank = Rank, Clinical_Care_Rank = Rank.3, Socioeconomic_Rank = Rank.4)
```r

 4.1.3 Isolating Racial Disparity Metrics (Files 5 & 6) 

File 5 (Ranked Measure Data) had over 100 columns, including raw rates, confidence intervals, and data reliability flags. To keep the focus on racial equity, I used the contains() function to target the main Years of Potential Life Lost (YPLL) rates for Black and White populations while discarding secondary statistics. 

```{r clean_2021_file5}
 # Processing File 5: Targeted extraction of YPLL and Poor Health metrics clean21_rankedmeasures <- read.csv("data/raw/2021/Ranked Measure Data.csv", skip = 1) %>% dplyr::select(FIPS, YPLL_Rate = contains("Potential.Life.Lost.Rate"), YPLL_Black = contains("YPLL.Rate..Black"), YPLL_White = contains("YPLL.Rate..White"), Percent_Poor_Health = contains("Fair.or.Poor.Health")) %>% filter(!is.na(YPLL_Rate)) 

```r

4.1.4 Relational Integration and Data Type Audit 

The last step of the 2021 method was a multi-stage Relational Join using the Federal Information Processing Standard (FIPS) codes as the main key. After the join, I did a Data Type Audit. Because raw CSVs often import numeric rankings as character strings due to alphanumeric placeholders like "NR" (Not Reported), I converted these to numeric doubles. 


# Synthesizing the 2021 Master Dataset 

```{r join_and_audit_2021}

master_2021 <- clean21_measures %>% inner_join(clean21_ranked, by = "FIPS") %>% inner_join(clean21_subranked, by = "FIPS") %>% inner_join(clean21_rankedmeasures, by = "FIPS") # Auditing and forcing numeric types for mathematical validity master_2021$Median.Household.Income <- as.numeric(as.character(master_2021$Median.Household.Income)) master_2021$Health_Factors_Rank <- as.numeric(as.character(master_2021$Health_Factors_Rank)) master_2021$YPLL_Black <- as.numeric(as.character(master_2021$YPLL_Black)) master_2021$YPLL_White <- as.numeric(as.character(master_2021$YPLL_White)) # Listwise deletion of incomplete records master_2021 <- na.omit(master_2021) # Final Verification print(paste("Final 2021 County Count:", nrow(master_2021))) str(master_2021[c("Median.Household.Income", "YPLL_Black")])
```r
---
title: "CSC 308 Final Project - Individual Contribution"
author: "Emmanuella Obidike"
date: "April 2026"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
knitr::opts_chunk$set(echo = TRUE)

# Load dataset and filter selected states
master_2021 <- read.csv("Master_2021_Cleaned.csv")
data21 <- master_2021
```

# Project Overview

This document represents my individual contribution to the group project titled Exploring Healthcare Access Disparities Through Data-Driven Visualization and Analysis. My work focuses on exploratory data analysis (EDA) and interpretation using the cleaned 2021 dataset.

The goal of this analysis is to examine relationships between income, healthcare access, and health outcomes across U.S. counties. By analyzing these patterns, we can better understand how socioeconomic conditions influence healthcare quality and overall population health.

---

# Individual Contribution - Emmanuella Obidike

**Role:** Exploratory Data Analysis (EDA), Statistical Analysis, and Interpretation

For this project, I analyzed the cleaned 2021 dataset to identify patterns between income levels, clinical care access, and health outcomes. My work included generating summary statistics, creating visualizations, performing correlation analysis, and interpreting key findings.

All code and analysis presented in this document reflect my individual contribution.

---

# Dataset Description

The dataset used for this analysis is Master_2021_Cleaned.csv, which contains county-level data across the United States, including:

* Median household income
* Clinical care rankings
* Health factors and health outcomes rankings
* Demographic and health-related indicators

The analysis uses the full dataset (all U.S. counties) to provide a nationwide perspective on healthcare disparities.

---

# Summary Statistics

```{r}
summary_table <- data21 %>%
  summarise(
    avg_income = mean(Median.Household.Income, na.rm = TRUE),
    avg_clinical_rank = mean(Clinical_Care_Rank, na.rm = TRUE)
  )

summary_table
```

This summary provides an overall average of income and clinical care rankings across U.S. counties.

---

# Visualization 1: Income vs Clinical Care

```{r}
ggplot(data21, aes(x = Median.Household.Income, y = Clinical_Care_Rank)) +
  geom_point(color = "blue", alpha = 0.3) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    title = "Income vs Clinical Care Across U.S. Counties (2021)",
    x = "Median Household Income",
    y = "Clinical Care Rank (Lower = Better)"
  ) +
  theme_minimal()
```

**Interpretation**

This plot shows a negative relationship between income and clinical care rankings. Counties with higher median household income tend to have better (lower) clinical care rankings, suggesting that economic resources play a role in access to healthcare services.

---

# Correlation: Income vs Clinical Care

```{r}
correlation <- cor(
  data21$Median.Household.Income,
  data21$Clinical_Care_Rank,
  use = "complete.obs"
)

correlation
```

**Interpretation**

The correlation is negative, indicating that as income increases, clinical care rankings improve. This supports the idea that income is an important factor in healthcare access.

---

# Visualization 2: Health Factors vs Health Outcomes

```{r}
ggplot(data21, aes(x = Health_Factors_Rank, y = Health_Outcomes_Rank)) +
  geom_point(alpha = 0.3, color = "darkgreen") +
  labs(
    title = "Health Factors vs Health Outcomes Across U.S. Counties (2021)",
    x = "Health Factors Rank (Lower = Better)",
    y = "Health Outcomes Rank (Lower = Better)"
  ) +
  theme_minimal()
```

**Interpretation**

There is a strong positive relationship between health factors and health outcomes. Counties with worse health factors tend to also have worse health outcomes, showing how environmental and behavioral conditions directly impact population health.

---

# Correlation: Health Factors vs Outcomes

```{r}
correlation2 <- cor(
  data21$Health_Factors_Rank,
  data21$Health_Outcomes_Rank,
  use = "complete.obs"
)

correlation2
```


**Interpretation**

The correlation is very strong, confirming that health factors are closely linked to health outcomes across counties.

---

# Visualization 3: Income Distribution

```{r}
ggplot(data21, aes(x = Median.Household.Income)) +
  geom_histogram(fill = "purple", bins = 30, alpha = 0.7) +
  labs(
    title = "Distribution of Median Household Income (U.S. Counties, 2021)",
    x = "Median Household Income",
    y = "Count"
  ) +
  theme_minimal()
```


**Interpretation**

Most counties fall within a middle-income range, with fewer counties at higher income levels, indicating a right-skewed distribution.

---

# Visualization 4: Racial Disparities (YPLL)

```{r}
racial_data <- data21 %>%
  select(YPLL_Black, YPLL_White) %>%
  pivot_longer(cols = everything(), names_to = "Group", values_to = "YPLL")

ggplot(racial_data, aes(x = Group, y = YPLL, fill = Group)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Racial Disparities in Premature Mortality (YPLL)",
    x = "Population Group",
    y = "YPLL Rate"
  ) +
  theme_minimal()
```

**Interpretation**

The plot shows noticeable differences in premature mortality between groups, suggesting disparities in health outcomes across populations.

---

# Visualization 5: Income vs Health Outcomes

```{r}
ggplot(data21, aes(x = Median.Household.Income, y = Health_Outcomes_Rank)) +
  geom_point(alpha = 0.3, color = "darkred") +
  labs(
    title = "Income vs Health Outcomes Across U.S. Counties (2021)",
    x = "Median Household Income",
    y = "Health Outcomes Rank"
  ) +
  theme_minimal()
```

**Interpretation**

Higher income is generally associated with better health outcomes, though there is still variability across counties.

---

# Visualization 6: Clinical Care Distribution

```{r}
ggplot(data21, aes(x = Clinical_Care_Rank)) +
  geom_histogram(fill = "blue", bins = 30, alpha = 0.7) +
  labs(
    title = "Distribution of Clinical Care Rankings",
    x = "Clinical Care Rank",
    y = "Count"
  ) +
  theme_minimal()
```

**Interpretation**

Clinical care rankings are concentrated at lower values (better rankings), with fewer counties having poor rankings.

---

# Conclusion

The analysis reveals clear relationships between socioeconomic conditions and healthcare access.

Higher income levels are associated with better clinical care rankings, suggesting that financial resources influence access to healthcare services. Additionally, the strong relationship between health factors and health outcomes shows that environmental and behavioral conditions play a major role in overall population health.

These patterns highlight broader disparities in healthcare access across U.S. counties

Overall, this analysis demonstrates how exploratory data analysis can transform complex datasets into meaningful insights about public health disparities.

---

# Final Statement

All code and analysis in this document reflect my individual contribution to the project.

---
title: "CSC 308 Final Project - Individual Portfolio"
author: "Chloe"
date: "April 2026"
output: html_document
---
  
  ```
{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
```

# Introduction

For this project, my role focused on **discussing the implications of the results and explaining why the findings matter**.  
Instead of mainly creating the exploratory analysis, my contribution centered on interpreting the significance of the patterns found in the data and connecting them to real-world healthcare access issues.

The purpose of my analysis was to explain how the project’s findings relate to healthcare inequality, differences across communities, and the broader importance of socioeconomic and health-related conditions.

---
  
  # Dataset and Preparation
  
  I used the cleaned 2021 project dataset:
  
  `processed_data/2021/Master_2021_Cleaned.csv`

The analysis focused on the following states:
  
  - Virginia  
- North Carolina  
- Maryland  
- Georgia  
- Florida  

```{r}
data21 <- read.csv("processed_data/2021/Master_2021_Cleaned.csv")

states <- c("Virginia", "North Carolina", "Maryland", "Georgia", "Florida")

filtered_data <- data21 %>%
  filter(State %in% states)

head(filtered_data)
```

---
  
  # Summary Statistics
  
  I first reviewed summary statistics to support discussion of the project findings.

```{r}
summary_table <- filtered_data %>%
  group_by(State) %>%
  summarise(
    avg_income = mean(Median.Household.Income, na.rm = TRUE),
    avg_clinical_rank = mean(Clinical_Care_Rank, na.rm = TRUE)
  )

summary_table
```

## Discussion

This summary helps show that the selected states differ in both income and clinical care rankings.  
These differences matter because they suggest that economic conditions may shape the quality of healthcare access available in different communities.

The results also show that healthcare disparities are not evenly distributed. Some states appear to have stronger average conditions than others, which may reflect broader differences in resources, provider availability, and community health support.

---
  
  # Visualization: Income vs Clinical Care
  
  ```{r}
ggplot(filtered_data, aes(x = Median.Household.Income, y = Clinical_Care_Rank)) +
  geom_point(alpha = 0.4) +
  labs(
    title = "Income vs Clinical Care (2021)",
    x = "Median Household Income",
    y = "Clinical Care Rank"
  ) +
  theme_minimal()
```

## Discussion

This plot suggests that counties with higher household income tend to have better clinical care rankings.  
Because lower ranking values indicate better clinical care, the pattern implies that wealthier counties may have advantages in healthcare access and quality.

This has important implications. It suggests that healthcare access is connected to economic inequality. Communities with fewer financial resources may also face greater difficulty accessing providers, services, and other forms of care.

In a real-world sense, this means healthcare differences are not only medical issues. They are also social and economic issues.

---
  
  # Correlation: Income vs Clinical Care
  
  ```{r}
correlation <- cor(
  filtered_data$Median.Household.Income,
  filtered_data$Clinical_Care_Rank,
  use = "complete.obs"
)

correlation
```

## Discussion

The correlation is negative, which supports the idea that higher income is associated with better clinical care rankings.

This result matters because it strengthens the visual pattern shown in the scatter plot.  
It suggests that income may play a meaningful role in shaping healthcare access across counties.

The implication of this finding is that improving healthcare equity may require more than expanding medical services alone. It may also require attention to larger economic inequalities.

---
  
  # Visualization: Average Income by State
  
  ```{r}
ggplot(summary_table, aes(x = State, y = avg_income)) +
  geom_col() +
  labs(
    title = "Average Income by State (2021)",
    x = "State",
    y = "Average Income"
  ) +
  theme_minimal()
```

## Discussion

This state comparison shows that average income is not the same across the selected states.  
Some states have noticeably higher average income than others.

This is important because differences in income may help explain why healthcare rankings are also uneven.  
If income shapes access to insurance, transportation, providers, and preventive care, then economic differences may contribute to long-term health disparities across regions.

---
  
  # Correlation: Health Factors vs Health Outcomes
  
  ```{r}
correlation2 <- cor(
  filtered_data$Health_Factors_Rank,
  filtered_data$Health_Outcomes_Rank,
  use = "complete.obs"
)

correlation2
```

## Discussion

The relationship between health factors and health outcomes is very strong.  
This means counties with weaker health-related conditions also tend to have poorer overall health outcomes.

This finding has a broader implication for public health. It suggests that healthcare outcomes are shaped by more than treatment alone. Social and environmental conditions also matter.

Because of this, policies that focus only on hospitals or medical care may not be enough. Improving health outcomes may also require investments in the larger conditions that affect people’s daily lives.

---
  
  # Real-World Implications
  
  The results of this project suggest several important implications:
  
  - Economic inequality may contribute to unequal healthcare access  
- Better clinical care is often connected to stronger financial conditions  
- Health outcomes are influenced by broader health factors, not only medical care  
- Communities with weaker support conditions may face long-term disadvantages in health and healthcare access  

These findings matter because they show that healthcare disparities are part of a larger structural issue.  
This means that solving healthcare inequality likely requires both medical improvements and attention to larger social and economic conditions.

---
  
  # Limitations
  
  Although the results are meaningful, there are still some limitations.

First, rankings and summary values do not capture every detail about the healthcare experience of individuals.  
Second, correlation does not prove causation. Even though the relationships are strong, the data does not prove that one variable directly causes the other.  
Finally, the analysis focuses on selected states and one cleaned dataset, so the findings should be interpreted within that scope.

---
  
  # Conclusion
  
  My contribution to this project focused on discussing the meaning and significance of the results.  
The analysis suggests that differences in income, clinical care, and health-related conditions are closely connected.

Overall, these results show that healthcare access disparities are tied to larger patterns of inequality.  
This makes the findings important not only for data analysis, but also for understanding real-world healthcare challenges.

---
  
  # Individual Contribution
  
  For this project, I was responsible for:
  
  - Explaining the significance of the project results  
- Discussing the implications of relationships found in the data  
- Connecting findings to real-world healthcare inequality  
- Supporting discussion with summary statistics, visualizations, and correlations  
- Reflecting on limitations and broader meaning of the results  

All code in this document reflects my individual contribution to the project.

---
title: "CSC 308 Final Project - Individual Portfolio"
author: "Joshua"
date: "April 2026"
output: html_document
---
This visualization explores the relationship between income and healthcare access. The results will show that lower income levels are associated with higher uninsured rates which portray that economic factors play a vital role in healthcare accessibility.

This analysis supports the overall findings of the project because it shows how financial disparities influence health outcomes across different populations.

